Northern Pikeminnow Abundance in Deadwater Slough, Salmon River, Idaho, and Potential Impacts to Local Chinook Salmon Populations

### 2022-04-07 analysis/paper/paper.Rmd
near final edits of Discussion

Using hook-and-line angling, we caught `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "n_fish"])`, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "n_fish"])`, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "n_fish"])` Northern Pikeminnow during the fall 2019, fall 2020, and spring 2021 surveys, respectively, for a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` fish. The CPUE was `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2019", "catch_per_hr"], digits = 3)` for fall 2019, `r prettyNum(total_cpue[total_cpue$event_name == "Fall 2020", "catch_per_hr"], digits = 3)` for fall 2020, and `r prettyNum(total_cpue[total_cpue$event_name == "Spring 2021", "catch_per_hr"], digits = 2)` for spring 2021. During the mark-recapture events, there were `r as.character(english(sum(N_mods[[3]][[1]]$m)))` recaptures in fall 2019, including `r as.character(english(LP_mod[[3]][[1]]$m))` unique individuals and `r as.character(english(sum(N_mods[[3]][[3]]$m)))` recaptures in fall 2020, including `r as.character(english(LP_mod[[3]][[3]]$m))` unique individuals. Summaries of parameters for both the single census and multiple census mark-recapture estimators are provided in Table \@ref(tab:abund-data-lp) and Table \@ref(tab:abund-data-schn).

Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). We estimated larger populations of Northern Pikeminnow using the unadjusted multiple-census estimator compared to adjusted and single census estimators (Figure \@ref(fig:abund-fig)). Because our sampling design most closely matched a multiple census estimator and required a small sample size modification, we deemed the Schnabel estimates most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r prettyNum(round(mean(all_est$Lci[all_est$model == "Schnabel"])), big.mark = ",")` - `r prettyNum(round(mean(all_est$Uci[all_est$model == "Schnabel"])), big.mark = ",")`). Using the delayed-mixing Schnabel estimator, the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` stomachs. Northern Pikeminnow captured during spring 2021 had a higher proportion of fish content in their diet compared to individuals captured in the fall surveys (Table \@ref(tab:gastric-tab)). The mean wet weight of total contents for an individual was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; SD = `r round(sd(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants made up `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of all stomach contents examined.

@@

During the fall DSR emigration, we estimate an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consists of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet is fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` in the fall, `r prettyNum(round(spring_abund$N_S), big.mark = ",")` in the spring 2021, and a diet consisting of 60% fish, the Northern Pikeminnow population can consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

Given yearly fish consumption estimates during the fall DSR and spring NRR periods, we produced a sensitivity analysis to calculated consumption of juvenile Chinook Salmon where 30%, 50%, and 65% of total fish prey consumed are juvenile Chinook Salmon (Figure \@ref(fig:juv-con-p)). At the median value of 50%, we estimated that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$juv_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$juv_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) juvenile Chinook Salmon will be consumed. Using the median Granite-to-Granite SAR, we estimated that juvenile consumption is equivalent to `r prettyNum(round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$adult_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$adult_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) adults (Figure \@ref(fig:adult-imp-p)).

## Mark-Recapture Models

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupies Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledge that their methods were unsuitable for sampling offshore, noting that water velocity, depth, irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, at Deadwater Slough the maximum depth was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimated densities of Northern Pikeminnow suggest that slow-water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin; contemporary information on piscine predator abundances may be needed [@Widener2021].

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population, however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and unmarked fish. Given the size of the sampling area, and the short duration of our surveys, the closed population assumption is likely met. <!--We also caught one individual that was confirmed tagged in Deadwater Slough in a previous year suggesting that at least some individuals remain in the reach or at least migrate to Deadwater Slough multiple times, annually.-->This assumption can also affect results if mortality occurs for some marked fish released back to the population. During field processing, no mortalities or injury post release was observed; however, the potential for mortality amongst marked and released fish cannot be discounted. Next, our estimates assumed that capture and recapture events are random samples of the population. Angling methods may have a size selection bias, thereby limiting our abundance estimates to a size range susceptible to angling, resulting in a conservative estimate of the total population. We also question the validity of our assumption of equal catchability of individual fish between sampling events. During our study, we recaptured two individual fish during multiple days within a survey indicating that marked fish were still susceptible to angling.<!--We are also comfortable with the assumptions of no loss of marks or no mistaken or missed marks given that marks were physical (hole punches in fins) and that each fish was examined on a measuring board to collect lengths.--> However, we suspect that marked and released fish may not have mixed thoroughly back into the population prior to recapture events. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit "good" fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish. For example, if marked fish were only 70% as likely to be caught as unmarked fish, the true abundance is closer to 70% of our abundance estimate.

The spring abundance estimate was smaller than fall abundance estimates. Although our approach assumed equal capture probabilities between the fall and spring surveys, we believe higher spring flows may result in lower capture probability. Therefore, we believe our spring abundance estimate to be conservative.

In addition to the large population, the observed (`r round(psd)`%) PSD in this study was large compared to observations of 41% below Bonneville Dam and 18% in Bonneville Reservoir [@Winther2020]. This suggests that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. Notably, the Idaho state catch-and-release record Northern Pikeminnow, measuring `r max(gastric_df$length, na.rm = T)` mm TL, was caught in Deadwater Slough during the fall 2020 survey.

We performed gastric lavage on nearly all Northern Pikeminnow collected during this study. Observed prey species included juvenile Chinook Salmon as well as Redside Shiner *Richardsonius balteatus*, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*, though decomposition from digestion rendered most fish contents unidentifiable. Although our approach provided useful information on prey species, using angling and gastric lavage to quantify diet composition in Northern Pikeminnow can be problematic. First, diets of angled fish can differ from those collected by other methods such as electrofishing [@Hodgson1988]. For instance, angled fish are more likely to have empty stomachs [@Jurajda2016]; possibly because they are captured while actively searching for food, suggeting they are hungry. Second, unlike other predators such as Smallmouth Bass and Walleye, that are commonly sampled using gastric lavage [@Kamler2001], Northern Pikeminnow are cyprinids and therefore lack a true stomach. Consequently, performing gastric lavage on cyprinid species can be ineffective for accurately quantifying diet composition [@Hartleb1995], or alternately, requires a modification in methodology to flush food items out the vent of the fish [@Wasowicz1994]. Therefore, we consider our diet composition estimates and subsequent consumption estimates to be conservative.

## Bioenergetics

Our approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Generalizing energy densities using a single species is a common approach [@Petersen1999], however, we recognize that other salmonids including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. To address the prevalence of other fish in the Northern Pikeminnow's diet, we considered scenarios where DSR and NRR Chinook Salmon represent only 30%, 50%, and 65% of total fish prey consumed. These values may be considered conservative during the peak emigration periods when Chinook Salmon are abundant and the Northern Pikeminnow diet shifts almost entirely to piscivory, presumably on juvenile Chinook Salmon [@Shively1996; @Poe1991]. This is likely the case during the fall when DSR Chinook Salmon are the dominant prey species available in Deadwater Slough as 1) no (or at least minimal) hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent than Chinook Salmon emigration. During spring when natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, Sockeye Salmon) are present in the river, <50% Chinook Salmon in the Pikeminnow's diet may be more likely. Future work to quantify and identify the juvenile salmonid in Deadwater Slough during their seasonal migration would be useful to validate the diet composition assumptions used in our model.

## Potential Impacts to Salmon Populations

Recovery of Chinook Salmon in the Upper Salmon MPG is desired to support local fisheries and economies. At least five of the eight populations must meet criteria set forth by @McElhany2000 and the @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. We estimate that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` juvenile Chinook Salmon may be consumed by Northern Pikeminnow assuming that 60% of their diet is fish and 50% of fish prey are Chinook Salmon. Considering that diet scenarios used in our study are likely conservative assumptions, especially during the fall DSR emigration, the estimated number of Chinook Salmon consumed is substantial.

>>>>

Using hook-and-line angling, we caught a total of `r prettyNum(sum(total_cpue$n_fish), big.mark = ",")` Northern Pikeminnow over the course of the study including `r sum(N_mods[[3]][[1]]$m) + sum(N_mods[[3]][[3]]$m)` recaptures; the overall CPUE was `r round(sum(cpue_df$n_fish) / (sum(cpue_df$total_effort_min) / 60), 2)` Northern Pikeminnow per angler hour (table reference needed). Mark-recapture abundance estimates of Northern Pikeminnow in Deadwater Slough ranged from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2019"]), big.mark = ",")` in fall 2019 and from `r prettyNum(min(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` to `r prettyNum(max(all_est$N[all_est$event_name == "Fall 2020"]), big.mark = ",")` in fall 2020 (Table \@ref(tab:abund-tab)). We estimated larger populations of Northern Pikeminnow using the unadjusted multiple-census estimator compared to the single census and adjusted delayed-mixing multiple census estimators (Figure \@ref(fig:abund-fig)). Our sampling design most closely matched a multiple-census estimator; therefore, the Schnabel estimates were considered most appropriate. Accordingly, the mean Northern Pikeminnow abundance for the two fall sampling events was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel"]), big.mark = ",")` (95% CI: `r prettyNum(round(mean(all_est$Lci[all_est$model == "Schnabel"])), big.mark = ",")` - `r prettyNum(round(mean(all_est$Uci[all_est$model == "Schnabel"])), big.mark = ",")`) using the unadjusted Schabel estimator. Using the delayed-mixing Schnabel estimator the mean fall abundance estimate was `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` (95% CI: `r prettyNum(mean(all_est$Lci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` - `r prettyNum(mean(all_est$Uci[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")`). All subsequent analyses use results from the adjusted delayed-mixing Schnabel estimator. For spring 2021, we estimated `r prettyNum(round(spring_abund$N_S), big.mark = ",")` (95% CI: `r paste(prettyNum(round(c(spring_abund$Lci_S, spring_abund$Uci_S)), big.mark = ","), collapse = " - ")`) Northern Pikeminnow in Deadwater Slough. Those estimates translate to linear densities of `r prettyNum(round(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms), big.mark = ",")` and `r prettyNum(round(spring_abund$N_S / ddw_rkms), big.mark = ",")` Northern Pikeminnow per rkm and areal densities (fish/100 m$^2$) of `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_m2) * 100, 1)` and `r round((spring_abund$N_S / ddw_m2) * 100, 1)` for fall and spring, respectively.

@@

We found that gastric lavage successfully removed food items in all dissected individuals, indicating the method was effective. Of the `r prettyNum(gastric_tbl$lavaged[gastric_tbl$event_name == "Total"], big.mark = ",")` Northern Pikeminnow sampled using gastric lavage, we found contents in `r prettyNum(gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` (`r prettyNum((gastric_tbl$stmch_contents[gastric_tbl$event_name == "Total"] / gastric_tbl$lavaged[gastric_tbl$event_name == "Total"]) * 100, digits = 3)`%) stomachs and confirmed fish or fish parts in `r prettyNum(gastric_tbl$fish_contents[gastric_tbl$event_name == "Total"], big.mark = ",")` of those. Northern Pikeminnow captured during spring 2021 had a higher proportion of fish content in their diet compared to individuals captured in the fall surveys (Table \@ref(tab:gastric-tab)). The mean wet weight of total contents for an individual was `r prettyNum(mean(stmch_df$stomach_contents_weight), digits = 2)` g (median = `r prettyNum(median(stmch_df$stomach_contents_weight), digits = 3)` g; SD = `r round(sd(stmch_df$stomach_contents_weight), 2)` g). Overall, fish or fish remnants made up `r prettyNum(mean(stmch_df$p_fish) * 100, digits = 3)`% of all stomach contents examined.

@@

During the fall DSR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_fall_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f60_g"]` g of fish to maintain their body size, assuming 60% of their diet consists of fish prey (Figure \@ref(fig:bio-p)). Fall consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f30_g"]` g for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Fall" & bio_g_summ$name == "f90_g"]` g for a diet of 90% fish. During the spring NRR emigration, we estimated an average-size Northern Pikeminnow (`r round(avg_spring_NPM_TL, 1)` TL mm) to consume `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f60_g"]` g of fish, assuming 60% of their diet is fish prey (Figure \@ref(fig:bio-p)). Spring consumption ranged from `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f30_g"]` for a diet of 30% fish to `r bio_g_summ$cumm_g[bio_g_summ$season == "Spring" & bio_g_summ$name == "f90_g"]` for a diet of 90% fish. Daily consumption rates were higher early in the fall and late in the spring and corresponded with water temperatures above 8 degrees Celsius (Figure \@ref(fig:bio-p)). Given estimated mean Northern Pikeminnow population sizes of `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` in the fall and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` in the spring 2021, and a diet consisting of 60% fish, the Northern Pikeminnow population is estimated to consume `r round(tot_g_DSR, 1)` metric tons of fish during the fall DSR emigration and `r round(tot_g_NRR, 1)` metric tons during the spring NRR emigration in Deadwater Slough.

@@

Given yearly fish consumption estimates during the fall DSR and spring NRR periods, we produced a sensitivity analysis to calculate the consumption of juvenile Chinook Salmon where 30%, 50%, and 65% of total fish prey consumed are juvenile Chinook Salmon (Figure \@ref(fig:juv-con-p)). At the median value of 50%, we estimated that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$juv_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$juv_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) juvenile Chinook Salmon will be consumed. Using the median Granite-to-Granite SAR for Chinook Salmon in the Upper Salmon River, we estimated the "adult equivalents" of juveniles consumed to be `r prettyNum(round(total_impacts$adult_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` (95% CI: `r prettyNum(round(total_impacts$adult_Lci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` - `r prettyNum(round(total_impacts$adult_Uci[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")`) adults (Figure \@ref(fig:adult-imp-p)).

@@

All mark-recapture abundance estimators suggest a large population of Northern Pikeminnow occupies Deadwater Slough. The mean population size of Northern Pikeminnow was estimated to be `r prettyNum(mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]), big.mark = ",")` during the fall DSR emigration and `r prettyNum(round(spring_abund$N_S), big.mark = ",")` during the spring NRR emigration even after accounting for the potential of delayed mixing of marked fish. Those estimates translate to linear densities (fish/rkm) that are `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / john_day_dens)` and `r round((spring_abund$N_S / ddw_rkms) / john_day_dens)` times greater than densities reported by @Beamesderfer1991 for the John Day Reservoir and approximately `r round((mean(all_est$N[all_est$model == "Schnabel - Delayed Mixing"]) / ddw_rkms) / low_columbia_fish_km)` and `r round((spring_abund$N_S / ddw_rkms) / low_columbia_fish_km)` times greater than projections for the lower Columbia River [@Beamesderfer1996]. @Beamesderfer1991 acknowledged that their methods were unsuitable for sampling offshore, noting that water velocity, depth, and irregular bottom contours, and barge traffic made sampling offshore ineffective; therefore it is possible that their estimates may have been low. Conversely, at Deadwater Slough the maximum depth was approximately 6 m with a relatively homogenous bottom contour, making angling an effective method throughout the entire reach. Our estimated densities of Northern Pikeminnow suggest that slow-water reaches outside of the reservoir complexes on the Snake and Columbia rivers may support exceptionally high predator densities, consistent with findings of @Harnish2014, @Gray2001, and @Zimmerman1999. Given the dearth of current data available in the Columbia River basin, it is unclear how Northern Pikeminnow abundance estimates in this study compare to elsewhere in the Columbia River Basin; contemporary information on piscine predator abundances may be needed [@Widener2021].

Several assumptions in our estimators may have influenced the magnitude of our abundance results. First, all models assumed a closed population; however, it is possible that immigration and emigration occurred. Nevertheless, our estimators still provide unbiased estimates of abundance assuming the immigration and emigration rates are equal between marked and marked fish. Emigration of marked individuals would reduce the marking fraction in the population resulting in an upward bias of abundance estimate, but emigration rates would need to be substantial. Further, the multiple census estimators are more robust to this assumption because the marking fraction is estimated daily. We additionally have no reason to believe that emigration occurs during the two-week window as Deadwater Slough provides favorable habitat to Northern Pikeminnow relative to adjacent reaches. Given the size of the sampling area and the short duration of our surveys, the closed population assumption is likely met. This assumption can also affect results if mortality for some marked fish released back to the population. During field processing, no mortalities or injury post release was observed and fish injured prior to release were retained. Nevertheless, the potential for mortality amongst marked and released fish cannot be discounted.

Next, our abundance estimates assumed that capture and recapture events are random samples of the population. Angling methods often have a size selection bias, thereby limiting abundance estimates to a size range susceptible to angling and resulting in a conservative estimate of the total population. We also question the validity of our assumption of equal catchability of individual fish between sampling events. Although we recaptured two individual fish during multiple days within a survey (indicating that marked fish were still susceptible to angling), we suspect that marked and released fish may not have immediately mixed thoroughly back into the population. Although we attempted to release marked fish evenly throughout Deadwater Slough, we anecdotally observed anglers congregating in particular areas of the slough to socialize or exploit “good” fishing locales. Violations of the assumption of equal catchability between marked and unmarked fish may lead to an overestimate of abundance of Northern Pikeminnow in our study if marked fish have lower catchability. This bias will be proportional to the difference in catchability of marked and unmarked fish. For example, if marked fish were only 70% as likely to be caught as unmarked fish, the true abundance is closer to 70% of our abundance estimate. This would additionally account for marked and released fish being “hook shy” for a period of time. The adjusted delayed-mixing Schnabel estimator may account for these biases; however the magnitude of bias is unknown.

The spring abundance estimate was smaller than fall abundance estimates due to the spring CPUE being approximately half of the mean fall CPUE. Although our approach assumed equal capture probabilities between the fall and spring surveys, we believe that higher spring flows may result in lower capture probability of Northern Pikeminnow. Therefore, we believe our spring abundance estimate to be conservative.

In addition to the large population abundance, the observed (`r round(psd)`%) PSD in this study was greater than observations of 41% below Bonneville Dam and 18% in Bonneville Reservoir [@Winther2020]. This suggests that a larger fraction of Northern Pikeminnow in Deadwater Slough are of a quality size relative to populations reported elsewhere in the Columbia River. Notably, the Idaho state catch-and-release record Northern Pikeminnow, measuring `r max(gastric_df$length, na.rm = T)` mm TL, was caught in Deadwater Slough during the fall 2020 survey.

We performed gastric lavage on nearly all Northern Pikeminnow collected during this study. Observed prey species included juvenile Chinook Salmon as well as Redside Shiner *Richardsonius balteatus*, Largescale Sucker *Catostomus macrocheilus*, sculpin *Cottus* spp., and Mountain Whitefish *Prosopium williamsoni*, though decomposition from digestion rendered most fish contents unidentifiable. Although our approach provided useful information on prey species, the use of angling and gastric lavage to quantify diet composition in Northern Pikeminnow can be problematic. First, diets of angled fish can differ from those collected by other methods such as electrofishing [@Hodgson1988]. For instance, angled fish are more likely to have empty stomachs [@Jurajda2016]; possibly because they are captured while actively searching for food, suggesting they are hungry. Second, unlike other predators such as Smallmouth Bass and Walleye, that are commonly sampled using gastric lavage [@Kamler2001], Northern Pikeminnow are cyprinids and therefore lack a true stomach. Consequently, performing gastric lavage on cyprinid species can be ineffective for accurately quantifying diet composition [@Hartleb1995], or alternately, requires a modification in methodology to flush food items out the vent of the fish [@Wasowicz1994]. Therefore, we consider our diet composition estimates from gastric lavage to be conservative.

Our bioenergetics approach assumed that all available prey consumed by Northern Pikeminnow have energy densities equal to juvenile Chinook Salmon. Generalizing energy densities using a single species is a common approach [@Petersen1999]. Other salmonids including juvenile steelhead and Sockeye Salmon are likely also available prey in Deadwater Slough, especially during the spring migration, and so generalizing energy densities among salmonids may not be problematic. Additionally, hatchery smolt releases (Chinook Salmon, steelhead, and Sockeye Salmon) are also prevalent in the mainstem Salmon River during the spring migration. To address the prevalence of other fish in the Northern Pikeminnow’s diet, we considered scenarios where DSR and NRR Chinook Salmon represent only 30%, 50%, and 65% of total fish prey consumed. These values may be considered conservative during the peak emigration periods when Chinook Salmon are abundant and the Northern Pikeminnow diet shifts almost entirely to piscivory, presumably on juvenile Chinook Salmon [@Shively1996; @Poe1991]. This is likely the case during fall months when DSR Chinook Salmon are the dominant prey species available in Deadwater Slough as 1) few to no hatchery releases are present in the river and 2) fall emigrations of steelhead and Sockeye Salmon are less prominent than Chinook Salmon. During spring when natural-origin steelhead and Sockeye Salmon are also actively emigrating from the Upper Salmon River and hatchery-origin releases of all three species (Chinook Salmon, steelhead, Sockeye Salmon) are present in the river, less than 50% Chinook Salmon in the Northern Pikeminnow’s diet may be more likely. Predation on juveniles from any of the ESA-listed salmonid species is detrimental to their recovery. Future work to quantify and identify juvenile salmonids in Deadwater Slough during their seasonal migration would be useful to validate the diet composition assumptions used in our model and to understand impacts to local populations.

Chinook Salmon populations above Deadwater Slough are within the Upper Salmon major population group (MPG) which supports eight independent, extant populations including Salmon River (above Redfish Lake Creek), Valley Creek, Yankee Fork Salmon River, East Fork Salmon River, Salmon River (mainstem below Redfish Lake Creek), Pahsimeroi River, Lemhi River, and North Fork Salmon River [@NOAA2017]. Recovery of the MPG is desired to support local fisheries and economies. At least five of the eight populations must meet criteria set forth by @McElhany2000 and the @ICTRT2007 for the MPG to be considered viable and for recovery of the Snake River Evolutionary Significant Unit. We estimated that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50]), big.mark = ",")` juvenile Chinook Salmon may be consumed by Northern Pikeminnow assuming that 60% of their diet is fish and 50% of fish prey are Chinook Salmon. For context, this is approximately `r round(total_impacts$juv_N[total_impacts$perc_diet == 60 & total_impacts$p_chnk == 0.50] / 45416, 2)` times the average annual combined DSR and NRR emigration estimates for brood years 2011 - 2018 at a rotary screw trap located in the lower Lemhi River [@McClure2021; @Feeken2020; @Poole2019], the largest population in the Upper Salmon MPG. Even under the most conservative scenario where 30% of the Northern Pikeminnow diet is fish and 30% of fish prey are Chinook Salmon, we estimate that `r prettyNum(round(total_impacts$juv_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30]), big.mark = ",")` juvenile Chinook Salmon may be consumed which is `r round((total_impacts$juv_N[total_impacts$perc_diet == 30 & total_impacts$p_chnk == 0.30] / 45416) * 100)`% of the total DSR and NRR emigration for brood years 2011 - 2018 from the Lemhi River. Considering that diet scenarios used in our study are likely conservative assumptions, especially during the fall DSR emigration, the estimated number of Chinook Salmon consumed is substantial.
